What's in a name? Disambiguation strategies for relational descriptions

### 2023-05-14 2023 Disambiguation/Disambiguation.tex
Update changes

An invariant in this system is:

\[
\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}} \typesubset \mathrm{scheduled}_\mathrm{Event * Time}\typesemi\mathrm{before}_{\mathrm{Time * Time}}
\]

@@

In Ampersand's implementation, Ampersand does not require writing the types when writing Invariants.
We can consider expressing the same invariant as earlier without writing types: $\mathrm{before}\typesemi \mathrm{scheduled} \typesubset \mathrm{scheduled}\typesemi \mathrm{before}$.
If we require that invariants are well-typed, there is only one way of annotating such expressions with types.
The use of the possibly ambiguous relation name $\mathrm{before}$ is allowed as long as it is clear which relation is intended, as is the case here.
We refer to the problem of determining which relation is intended as the disambiguation problem.%, which we investigate more closely in a separate section.
%
%\subsection{Generalization}
%A common task in semantic modeling is to identify that some concepts are generalizations or specializations of other concepts.
%This section motivates the need for a special language feature to deal with generalizations.
%
%As an example, our planning application allows regular working hours to be set, which means that some times fall during regular hours.
%This means that such times would be a specialization of a time.
%We can model this with a relation called $\mathrm{worktime}$ and an invariant: $\mathrm{worktime} \subseteq \mathbb{1}_{Time}$, where $\mathbb{1}_{A} = \{(a,a) \mid a\in A \}$ is the relational identity.
%Another way to model this, is to introduce a concept $\mathrm{Worktime}$ and use $\mathbb{1}_\mathrm{Worktime}$ in the place of the relation $\mathrm{worktime}$, along with the invariant $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$.
%However, this latter invariant is ill-typed: the expression on the left of the subset ($\subseteq$) is of type $(\mathrm{Worktime},\mathrm{Worktime})$, while the expression on the right is of type $(\mathrm{Time},\mathrm{Time})$.
%To allow expressions like $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$ in a programming language, we need a more refined notion of which expressions are `type correct'.
%A paper by van der Woude and Joosten shows how one can do this~\cite{Woude11}.
%This paper has largely the same notion of type correctness.
%However, that paper does not consider the interplay with disambiguation which we consider here.
%
%\subsection{Two interpretations.}
%There are two different ways in which we can think of the two binary relations called $\mathrm{before}$.
%In one interpretation, we can say that $\mathrm{before}$ is a single binary relation, defined over the (disjoint) union between Task and Time.
%Whenever we use $\mathrm{before}$, we restrict its type to either \verb=Task * Task= or \verb=Time * Time= to get two completely separate behaviors.
%This means that we read $\mathrm{before}_{\mathrm{Task * Task}}$ as $\mathrm{before} \cap (\mathrm{Task}\times\mathrm{Task})$.
%
%Another interpretation is to say that there are two entirely separate relations, $\mathrm{before}_{\mathrm{Task * Task}}$ and $\mathrm{before}_{\mathrm{Time * Time}}$.
%In this interpretation, we can consider ${\mathrm{Task * Task}}$ as a part of the name of the relation.
%
%Note our interpretation does not matter for the behavior we get from well-typed invariants like the one we saw before.
%However, a problem arises if we add generalization.
%
%If we were to state - for some obscure reason - that every Task is a Time, in our first interpretation we would expect that $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$, since $(\mathrm{Task}\times\mathrm{Task}) \subseteq (\mathrm{Time}\times\mathrm{Time})$.
%In our second interpretation, we would not expect the behavior to change.
%
%Observe that we are free to add $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$ as an invariant under the second interpretation, mimicking the behavior of the first interpretation under the second interpretation.
%Therefore, we conclude that working with the second interpretation is most flexible and propose to follow this interpretation.

@@

A disambiguation problem is given by an expression $E$ using the operations $\cup$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible types of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
A solution to the disambiguation problem is a well-typed expression $S$ in which the annotations to relation names corresponds to an element in $\mathcal{t}(n)$, such that dropping the types from the expression yields $E$.

@@

% @Bas, deze paragraaf is wellicht leesbaarder als je gewoon de functie à la Haskell uitschrijft.
For the top-down step, let $d(e,t)$ denote the disambiguated expression $e$ with the type $t$.
For the top-level call of $d(e,t)$, let $e=E$ and $t\in p(E)$.
Then $d(n,t) = n_{[t]}$, $d(e_1 \cup e_2, t) = d(e_1,t) \cup d(e_2,t)$ and $d(e_1 ; e_2, (x,y)) = d(e_1, (x,z)) ; d(e_2, (z,y))$ where $z$ is picked such that $(x,z)\in p(e_1)$ and $(z,y)\in p(e_2)$.
Such a $z$ is guaranteed to exist by how $p$ is defined.

@@

The disambiguation problem and the corresponding algorithm can be extended over all operations in relation algebra.
It can even be used to disambiguate special relations like the identity relation.
Let's take $E_2 = \mathrm{scheduled};(\mathrm{before} \cup \mathbb{1})$ as an expression.
The types are as in the previous example, with $\mathcal{t}(\mathbb{1}) = \left\{(t,t)\mid t\in \mathcal{T}\right\}$.
This gives $p(E_2) = \{\mathrm{Task} * \mathrm{Time}\}$ and $d(\{\mathrm{Task} * \mathrm{Time}\},E_2) = \mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}};(\mathrm{before}_{\mathrm{Time} * \mathrm{Time}}\ \cup\ \mathbb{1}_{\mathrm{Time}})$, as desired.

\section{Solving Disambiguation with Generalization}
Before giving a solution to this % @Bas, een referentie over de grens van een paragraaf is taalkundig al niet fraai, maar over de grens van een section is niet slim. Als lezer wil ik iets meer duidelijkheid dan "this" als het gaat over welk probleem je naar verwijst.
problem, we first observe that the disambiguation algorithm discussed earlier fails for our running example.
Following the declarations of the script, we let $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Event}, \mathrm{Time})\}$.
Now the above algorithm would conclude with an error on $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.

@@

This paper follows a similar approach: We define a language to be used by the Ampersand programmer that allows writing relations without their types as well as subtyping.
Next, we define a language that is interpreted as a heterogeneous relation algebra with a typical set-interpretation.
% @Bas, de volgende zin snap ik niet...
The semantics of the language with subtyping will be determined by what the translation process gives us, but we validate the translation by inspecting certain examples and examining the validity of Tarski's axioms for this language under the resulting interpretation.
%We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a subtyping relation, so $\epsilon_{A*B}$ would exist only if $A$ is a subtype of $B$.
%In this paper, we relax that requirement.

@@

\paragraph{[Aside],} The Ampersand language also allows for \verb=CLASSIFY A IS B /\ C= statements, meaning that elements are in $A$ if and only if they are in $B$ and in $C$.
The information about this captured in $\preceq$ is just that $A \preceq B$ and $A\preceq C$, which means that information is lost.
We can consider the lost information when dealing with the $\typesemi$ operation by changing the $\epsilon$ relations introduced there.
However, it is an open problem to deal with this with the $\typeinter$ operation, so we give a simpler presentation that ignores this additional information in \verb=CLASSIFY _ IS _= statements for now.

>>>>

Note that `type' can refer to both a single concept, like \verb=Task=, \verb=Time= and \verb=Event=, and pairs thereof.
To avoid ambiguity, we refer to pairs of concepts (such as \verb=Task * Task=, \verb=Time * Time= and \verb=Even * Time=) as signatures.
We use `type' to refer to both.

Some invariants in this system are:

\begin{align}
\mathrm{before}_{\mathrm{Task * Task}}\typesemi\mathrm{scheduled}_{\mathrm{Event * Time}} &\typesubset \mathrm{scheduled}_\mathrm{Event * Time}\typesemi\mathrm{before}_{\mathrm{Time * Time}} \\
\mathrm{before}_{\mathrm{Task * Task}} \typeinter \typeident_\mathrm{Task} &\typesubset \typenil_\mathrm{Task * Task}
\end{align}

The intended meaning behind these invariants is: 1) The set of tasks that needs to be performed before some task $x$, all need to be scheduled at times before the time at which $x$ itself is scheduled. 2) It makes no sense to require a task to be performed before itself.

@@

In Ampersand's implementation, Ampersand does not require writing types when writing Invariants.
We can consider expressing the same invariant as earlier without writing types: $\mathrm{before}\typesemi \mathrm{scheduled} \typesubset \mathrm{scheduled}\typesemi \mathrm{before}$.
If we require that invariants are well-typed, there is only one possible intended meaning.
The use of the possibly ambiguous relation name $\mathrm{before}$ is allowed as long as it is clear which relation is intended, as is the case here.

For the other expression, we cannot omit writing all types: $\mathrm{before} \typeinter \typeident \typesubset \typenil$ could be interpreted as:
$\mathrm{before}_{\mathrm{Time * Time}} \typeinter \typeident_{\mathrm{Time}} \typesubset \typenil_{\mathrm{Time * Time}}$.
To avoid this issue, one might write \emph{some} type information: $\mathrm{before} \typeinter \typeident \typesubset \typenil_{\mathrm{Task * Task}}$.
From this, it should become clear which relations were intended (namely $\mathrm{before}_\mathrm{Task * Task}$ and $\typeident_\mathrm{Task}$), even though they appear elsewhere.

We refer to the problem of determining which relation is intended as the disambiguation problem.%, which we investigate more closely in a separate section.
%
%\subsection{Generalization}
%A common task in semantic modeling is to identify that some concepts are generalizations or specializations of other concepts.
%This section motivates the need for a special language feature to deal with generalizations.
%
%As an example, our planning application allows regular working hours to be set, which means that some times fall during regular hours.
%This means that such times would be a specialization of a time.
%We can model this with a relation called $\mathrm{worktime}$ and an invariant: $\mathrm{worktime} \subseteq \mathbb{1}_{Time}$, where $\mathbb{1}_{A} = \{(a,a) \mid a\in A \}$ is the relational identity.
%Another way to model this, is to introduce a concept $\mathrm{Worktime}$ and use $\mathbb{1}_\mathrm{Worktime}$ in the place of the relation $\mathrm{worktime}$, along with the invariant $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$.
%However, this latter invariant is ill-typed: the expression on the left of the subset ($\subseteq$) is of type $(\mathrm{Worktime},\mathrm{Worktime})$, while the expression on the right is of type $(\mathrm{Time},\mathrm{Time})$.
%To allow expressions like $\mathbb{1}_\mathrm{Worktime} \subseteq \mathbb{1}_{Time}$ in a programming language, we need a more refined notion of which expressions are `type correct'.
%A paper by van der Woude and Joosten shows how one can do this~\cite{Woude11}.
%This paper has largely the same notion of type correctness.
%However, that paper does not consider the interplay with disambiguation which we consider here.
%
%\subsection{Two interpretations.}
%There are two different ways in which we can think of the two binary relations called $\mathrm{before}$.
%In one interpretation, we can say that $\mathrm{before}$ is a single binary relation, defined over the (disjoint) union between Task and Time.
%Whenever we use $\mathrm{before}$, we restrict its type to either \verb=Task * Task= or \verb=Time * Time= to get two completely separate behaviors.
%This means that we read $\mathrm{before}_{\mathrm{Task * Task}}$ as $\mathrm{before} \cap (\mathrm{Task}\times\mathrm{Task})$.
%
%Another interpretation is to say that there are two entirely separate relations, $\mathrm{before}_{\mathrm{Task * Task}}$ and $\mathrm{before}_{\mathrm{Time * Time}}$.
%In this interpretation, we can consider ${\mathrm{Task * Task}}$ as a part of the name of the relation.
%
%Note our interpretation does not matter for the behavior we get from well-typed invariants like the one we saw before.
%However, a problem arises if we add generalization.
%
%If we were to state - for some obscure reason - that every Task is a Time, in our first interpretation we would expect that $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$, since $(\mathrm{Task}\times\mathrm{Task}) \subseteq (\mathrm{Time}\times\mathrm{Time})$.
%In our second interpretation, we would not expect the behavior to change.
%
%Observe that we are free to add $\mathrm{before}_{\mathrm{Task * Task}} = \mathrm{before}_{\mathrm{Time * Time}} \cap (\mathrm{Task}\times\mathrm{Task})$ as an invariant under the second interpretation, mimicking the behavior of the first interpretation under the second interpretation.
%Therefore, we conclude that working with the second interpretation is most flexible and propose to follow this interpretation.

@@

A disambiguation problem is given by an expression $E$ using the operations $\cup$ and $;$, and relation names from a set of names $\mathcal{N}$, and a set of possible signatures of each name: $\mathcal{t}(n) \subseteq \mathcal{T}\times \mathcal{T}$ for $n\in \mathcal{N}$.
A solution to the disambiguation problem is a well-typed expression $S$ in which the annotations to relation names corresponds to an element in $\mathcal{t}(n)$, such that dropping the types from the expression yields $E$.

@@

% @Bas, deze paragraaf is wellicht leesbaarder als je gewoon de functie à la Haskell uitschrijft.
% @Stef, zo staat het er al.
For the top-down step, let $d(e,t)$ denote the disambiguated expression $e$ with the type $t$.
For the top-level call of $d(e,t)$, let $e=E$ and $t\in p(E)$.
Then $d(n,t) = n_{[t]}$, $d(e_1 \cup e_2, t) = d(e_1,t) \cup d(e_2,t)$ and $d(e_1 ; e_2, (x,y)) = d(e_1, (x,z)) ; d(e_2, (z,y))$ where $z$ is picked such that $(x,z)\in p(e_1)$ and $(z,y)\in p(e_2)$.
Such a $z$ is guaranteed to exist by how $p$ is defined.

@@

The disambiguation problem and the corresponding algorithm can be extended over all operations in relation algebra.
It can even be used to disambiguate special relations like the identity relation.
Let's take $E_2 = \mathrm{scheduled};(\mathrm{before} \cup \mathbb{1})$ as an expression.
The signatures are as in the previous example, with $\mathcal{t}(\mathbb{1}) = \left\{(t,t)\mid t\in \mathcal{T}\right\}$.
This gives $p(E_2) = \{\mathrm{Task} * \mathrm{Time}\}$ and $d(\{\mathrm{Task} * \mathrm{Time}\},E_2) = \mathrm{scheduled}_{\mathrm{Task} * \mathrm{Time}};(\mathrm{before}_{\mathrm{Time} * \mathrm{Time}}\ \cup\ \mathbb{1}_{\mathrm{Time}})$, as desired.

\section{Solving Disambiguation with Generalization}
We first observe that the disambiguation algorithm discussed above fails for our running example.
Following the declarations of the script, we let $\mathcal{t}(\mathrm{before}) = \{(\mathrm{Task}, \mathrm{Task}),(\mathrm{Time}, \mathrm{Time})\}$ and $\mathcal{t}(\mathrm{scheduled}) = \{(\mathrm{Event}, \mathrm{Time})\}$.
Now the apply the algorithm on $E = (\mathrm{before};\mathrm{scheduled}) \cup (\mathrm{scheduled};\mathrm{before})$.
Since the disambiguation algorithm does not consider sub-typing, it would conclude there are is no way to well-type this expression.
In this section we address this shortcoming.

@@

%This paper follows a similar approach: We define a language to be used by the Ampersand programmer that allows writing relations without their types as well as subtyping.
%Next, we define a language that is interpreted as a heterogeneous relation algebra with a typical set-interpretation.
%% @Bas, de volgende zin snap ik niet...
%The semantics of the language with subtyping will be determined by what the translation process gives us, but we validate the translation by inspecting certain examples and examining the validity of Tarski's axioms for this language under the resulting interpretation.
%%We deviate slightly from the ideas by van der Woude and Joosten: Their $\epsilon$ relation is a subtyping relation, so $\epsilon_{A*B}$ would exist only if $A$ is a subtype of $B$.
%%In this paper, we relax that requirement.

@@

%\paragraph{[Aside]} The Ampersand language also allows for \verb=CLASSIFY A IS B /\ C= statements, meaning that elements are in $A$ if and only if they are in $B$ and in $C$.
%The information about this captured in $\preceq$ is just that $A \preceq B$ and $A\preceq C$, which means that information is lost.
%We can consider the lost information when dealing with the $\typesemi$ operation by changing the $\epsilon$ relations introduced there.
%However, it is an open problem to deal with this with the $\typeinter$ operation, so we give a simpler presentation that ignores this additional information in \verb=CLASSIFY _ IS _= statements for now.

Translating a term from $\mathcal{D}$ to $\mathcal{R}$ can be thought of as a search problem:
For each $\Xi$-term in $\mathcal{D}$, we can substitute it for an element in $\Xi$ to get a term in $\mathcal{R}$.
Our choice should satisfy the following conditions:
\begin{itemize}
\item The chosen term should be well-typed.
\item Depending on the choices that remain after applying these criteria, we need to decide to pick an arbitrary one or raise an error to the user.
\item We prefer some terms over others. In particular, we prefer to use $\epsilon_{A*B}$ relations where $A = B$ over those where $A\neq B$.
\end{itemize}

We illustrate these conditions one by one.
The first condition is per design: we aim to disambiguate terms by using that they are well-typed.
The second and third condition are best illustrated by giving an example of what could go wrong without them.

To illustrate why we prefer to use $\varepsilon_{A*B}$ relations where $A=B$, consider the expression: $(\mathrm{before} \typeinter \typeident) \typeunion \typetyped{\typenil}{\mathrm{Task * Task}}$. This expression is similar to one of the terms from our running example, but written as a proper $\mathcal{S}$-term.
Possible disambiguated terms include:
\begin{itemize}
\item $\varepsilon_T;(\varepsilon_T;\mathrm{before}_T;\varepsilon_T \cap \varepsilon_T;\typeident_\mathrm{Task};\varepsilon_T);\varepsilon_T \cup \varepsilon_T;F_T;\varepsilon_T$ with $T = \mathrm{Task*Task}$
\item
\item
\end{itemize}

the term $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typevee \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ with $\mathrm{Task}\preceq\mathrm{Event}$ would not disambiguate to the obvious term, since $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typetyped{\typevee}{\mathrm{Task}*\mathrm{Task}} \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ is also well-typed.
To deal with this, we prefer terms in which the $\epsilon$ relations have the same source and target.
